---
title: "age-length"
author: "Luke Hayden"
date: "August 1, 2018"
output: html_document
---

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}


library(dplyr)
library(ggbiplot)
library(tibble)
library(tidyr)
library(caret)

library(gtools)
```



```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

setwd("~/Documents/nstringjul18")
load(file="allns_data.rdata")
zklen <- read.csv("zk-age-len.csv")[2:3] %>%
  mutate(prep="ZK")

lena <- sampleinfo %>%
  filter(!is.na(wkage)) %>%
  select(length, wkage, prep) %>%
  rbind(zklen)


```

#Age-length relationship
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}



ggplot(lena, aes(y=length, x=wkage))+
  geom_point(aes(colour=prep))+
  geom_smooth()+
  theme_bw()+
  scale_colour_brewer(palette="Set1")+
  ggtitle("Age-length relationship in female Parhyale")+
  ylab("Body length (mm)")+
  xlab("Age (weeks)")
  
```
#Model length vs age

We will use the following models:

linear regression: age ~ length

linear regression: age ~ log(length)

linear regression: age ~ length^2

random forest (probably a sledgehammer to crack a nut)

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

lena$wkage <- as.numeric(lena$wkage)

lena$loglen <- log(lena$length)
lena$lensq <- (lena$length)^2



linmodel <- train(
  wkage~length, lena,
   method = "lm",
  trControl = trainControl(
    method = "cv", number = 5,
    repeats = 5, verboseIter = TRUE
  )
)

linmodel

lena$linpred <- predict(linmodel)

lena$linresid <- lena$wkage -lena$linpred


logmodel <- train(
  wkage~loglen, lena,
   method = "lm",
  trControl = trainControl(
    method = "cv", number = 5,
    repeats = 5, verboseIter = TRUE
  )
)

logmodel

lena$logpred <- predict(logmodel)

lena$logresid <-lena$wkage -lena$logpred

sqmodel <- train(
  wkage~lensq, lena,
   method = "lm",
  trControl = trainControl(
    method = "cv", number = 5,
    repeats = 5, verboseIter = TRUE
  )
)

sqmodel

lena$sqpred <- predict(sqmodel)
lena$sqresid <- lena$wkage -lena$sqpred


rfmodel <- train(
  wkage~length, lena,
   method = "rf",
  trControl = trainControl(
    method = "cv", number = 5,
    repeats = 5, verboseIter = TRUE
  )
)

rfmodel

lena$rfpred <- predict(rfmodel)
lena$rfresid <- lena$wkage -lena$rfpred

```

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

mlena <- gather(lena,key="model", value="pred", c("sqpred", "linpred", "logpred", "rfpred"))



ggplot(mlena, aes(x=wkage, y=pred, colour=model))+
  geom_point()+
  facet_wrap(~model, ncol=2)+
  theme_bw()+
  scale_colour_brewer(palette="Set1")+
  ggtitle("Model predicted vs real values", subtitle="3 types of regression + random forest")+
  xlab("Body length (mm)")+
  ylab("Age (weeks)")





#mlena <- gather(lena,key="model", value="pred", c("sqpred", "linpred", "logpred", "wkage"))
ggplot(mlena, aes(y=length, x=pred, colour=model))+
  geom_point()+
  geom_point(aes(x=wkage), colour="black")+
  facet_wrap(~model, ncol=2)+
  theme_bw()+
  scale_colour_brewer(palette="Set1")+
  ggtitle("Predictions of models: length vs age", subtitle="3 types of regression + random forest")+
  xlab("Body length (mm)")+
  ylab("Age (weeks)")
  


mlena <- gather(lena,key="model", value="resid", c("sqresid", "linresid", "logresid", "rfresid"))
ggplot(mlena, aes(x=wkage, y=resid, colour=model))+
  geom_bar(stat='identity')+
  facet_wrap(~model, ncol=1)+
  theme_bw()+
  scale_colour_brewer(palette="Set1")+
  ggtitle("Model residuals", subtitle="3 types of regression + random forest")+
  ylab("Residual")+
  xlab("Age (weeks)")

```

The best model is the linear regression of the squared length. With this, we get a very nice predictive power: `sqmodel$results`  Let's use this to predict the age of all samples where we only have length and not age. 


```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
sampleinfo$lensq <-sampleinfo$length^2
sampleinfo$predage <- NA

sampleinfo$predage[!is.na(sampleinfo$lensq)] <- predict(sqmodel, newdata=sampleinfo)

```




```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}


save(sampleinfo,ctall.norm,ctallgood.norm, codeset,geneinf,ctall.unnorm, ctall.intonly , file="allns_data.rdata")


```

