---
title: "Aging effect"
author: "Luke Hayden"
date: "August 3, 2018"
output: html_document
---
#Set-up and data import


```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(dplyr)
library(ggbiplot)
library(tibble)
library(tidyr)
library(caret)
library(RColorBrewer)
library(ggrepel)
library(gtools)
````




```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
setwd("~/Documents/nstringjul18")
load(file="allns_data.rdata")

````



#Old vs Young separation

Old vs young as main groups

##PCA old vs young

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}



log <- sampleinfo$reg == "" & 
  !(sampleinfo$age %in% c("N")) & 
  sampleinfo$sex == "F"&
  sampleinfo$exp %in% c("OvY", "Size-age") & 
  !(sampleinfo$sample %in% c("T","C", "U", "LFPD", "O1a", "SFPA", "LFIB","LFIA","AF", "Coh8")) &
 sampleinfo$over10reads > 100 &
  sampleinfo$ctg > 2 &
  sampleinfo$prep == "Luke"

bcmin <- as.matrix(ctallgood.norm[,log== TRUE])
bcmin.groups <- subset(sampleinfo$type, log==TRUE)

tbc <- as.matrix(t(bcmin))

bc.pca <- prcomp(tbc,center = TRUE,scale. = TRUE) 
bc.groups <- bcmin.groups


ggbiplot(bc.pca, obs.scale = 1, var.scale = 1, ellipse = FALSE, circle = TRUE, var.axes=F, labels=rownames(tbc), groups=bc.groups)+ggtitle("PCA of Old vs Young (females): 50 chosen markers")+
    theme_minimal() +
#  scale_colour_manual(values=c("cornflower blue", "red3", "black"))+
  theme(plot.title=element_text(size=8,face="bold"))+
#  geom_label(label=rownames(tbc), aes(colour=bc.groups))+
  scale_colour_manual(values= c("darkviolet","cornflower blue",  "red2"))


````

##What markers best separate old vs young?

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

md <- as.data.frame(subset(t(ctall.norm), sampleinfo$sex == "F" &
  #              sampleinfo$qual == "ok"& 
#                 sampleinfo$exp %in% c("OvY", "Size-age", "cohorts") & 
                 sampleinfo$ctg>2 & 
                sampleinfo$codeset == "phaw_1" &
                 sampleinfo$type %in% c("O", "Y", "M")))
 # !(sampleinfo$sample %in% c("T","C", "U", "LFPD", "O1a", "SFPA", "LFIB","LFIA","AF", "Coh8"))))



markerlist <- colnames(md)

md$sample <- rownames(md)


md <- left_join(md, select(sampleinfo, sample, predage), by="sample") %>%
  filter(!(is.na(predage))) %>%
  column_to_rownames(var="sample")



lmlist <- list()
arsqlist <- list()
rselist <- list()
plist <- list()
rmselist <- list()


for (i in 1:length(markerlist)){
  ilm <- lm(as.formula(paste0("predage~", markerlist[i])), md)
  lmlist[i] <- ilm
  arsqlist[i]<- summary(ilm)$adj.r.squared
  rmselist[i] <- sqrt(sum(residuals(ilm)^2) / df.residual(ilm))
  
  
}


lmdf <- data.frame(name =markerlist, rsq=unlist(arsqlist), rmse=unlist(rmselist)) %>%
  left_join(geneinf, by="name")


ggplot(lmdf, aes(x=name, y=rsq, fill=in61, label=name))+
  geom_bar(stat="identity")




````



```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

md <- t(subset(t(ctall.norm), sampleinfo$sex == "F" &
               sampleinfo$qual == "ok"& 
 #                sampleinfo$exp %in% c("OvY", "Size-age", "cohorts") & 
                 sampleinfo$ctg>2 & 
   sampleinfo$ctg <5 &
                sampleinfo$codeset == "phaw_1" &
                 sampleinfo$type %in% c("O", "Y", "M")&
 !(sampleinfo$sample %in% c("T","C", "U", "LFPD", "O1a", "SFPA", "LFIB","LFIA","AF", "Coh8"))
  
  
  ))

si <- subset(sampleinfo, sampleinfo$sample %in% colnames(md))
old <- t(subset(t(md), si$age == "O"))
young <- t(subset(t(md), si$age == "Y"))
#regen <- t(subset(t(ctg1to9), sampleinfo$type == "OR" & sampleinfo$qual != "bad"))



oy <- cbind(old, young)
t.result <-  apply(oy, 1, function (x) t.test(x[1:ncol(old)],x[ncol(old)+1: ncol(oy)]))


pvals <- data.frame(name=rownames(md), pvalovy = unlist(lapply(t.result, function(x) x$p.value)))



lmdf2 <- left_join(lmdf, pvals, by="name")
pthresh <- 0.05

ggplot(lmdf2, aes(x=pvalovy, y=rsq, fill=in61, label=name))+
  geom_point(shape=21)+
  scale_x_log10()+
  geom_vline(x=pthresh, xintercept = pthresh, linetype=2)+
  ggtitle(paste(length(lmdf$name[
    #lmdf2$rsq>0.05 & 
      lmdf2$pvalovy <pthresh])))#+geom_text()



```


```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

md <- t(subset(t(ctall.norm), sampleinfo$sex == "F" &
  #              sampleinfo$qual == "ok"& 
                 sampleinfo$exp %in% c("OvY", "Size-age", "cohorts") & 
                 sampleinfo$ctg>2 & 
                sampleinfo$codeset == "phaw_1" &
                 sampleinfo$type %in% c("O", "Y", "M")& 
  !(sampleinfo$sample %in% c("T","C", "U", "LFPD", "O1a", "SFPA", "LFIB","LFIA","AF", "Coh8"))))

si <- subset(sampleinfo, sampleinfo$sample %in% colnames(md))
old <- t(subset(t(md), si$age == "O"))
young <- t(subset(t(md), si$age == "Y"))
#regen <- t(subset(t(ctg1to9), sampleinfo$type == "OR" & sampleinfo$qual != "bad"))





nmd <- as.data.frame(md/rowMeans(md))


ns <- colnames(md)

nmd$name <- rownames(md)
nmd$sname <- substr(nmd$name, 1,2)

oy <- cbind(old, young)
t.result <-  apply(oy, 1, function (x) t.test(x[1:ncol(old)],x[ncol(old)+1: ncol(oy)]))



nmd$p_value <- unlist(lapply(t.result, function(x) x$p.value))

nmd <- left_join(nmd, geneinf, by="name")






nmd$YvOl2fc <- foldchange2logratio(foldchange(rowMeans(young), rowMeans(old)))


nmd$sname <- factor(nmd$sname, levels=nmd$sname[order(nmd$YvOl2fc)])
nmd$good <- nmd$p_value <0.05
nmd$di <- paste0(ifelse(nmd$dir == "upregulated with age", "up", "down"), ": ", nmd$sname)


#nmd <- as.data.frame(md/rowMeans(md))



````



##Expression/length relationship

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
mnmd <- gather(nmd, key="sample", value="normexp", ns)


mnmd <- left_join(mnmd, sampleinfo, by="sample")



ggplot(mnmd, aes(x=length, y=normexp))+
  geom_point(aes(fill=type), shape=21, colour="white")+
  geom_smooth(method="lm", aes(colour=dir), alpha=0.3, se=F, size=1)+
  theme_bw()+
 scale_y_continuous(trans="log2")+
  scale_colour_manual(values=c("orangered","forest green"))+
  scale_fill_manual(values=c("darkviolet", "cornflower blue",  "red3"))+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ylab("Normalised Expression") +
  xlab('Body length')+
  facet_wrap(~sname, scales="free_y")




ggplot(mnmd, aes(x=predage, y=normexp))+
  geom_point(aes(fill=type), shape=21, colour="white")+
  geom_smooth(method="lm", aes(colour=in61), alpha=0.3, se=F, size=1)+
  theme_bw()+
 scale_y_continuous(trans="log2")+
  scale_colour_manual(values=c("orangered","forest green"))+
  scale_fill_manual(values=c("darkviolet", "cornflower blue",  "red3"))+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ylab("Normalised Expression") +
  xlab('Body length')+
  facet_wrap(~dir)

  facet_wrap(~sname, scales="free_y")




ggplot(mnmd, aes(x=predage, y=normexp))+
  geom_point(aes(colour=type))+
#  geom_text(aes(label=sample, colour=type))+
  geom_smooth(method="lm", se=F, aes(size=sname),alpha=0.3,linetype=1, colour="gray30")+
  scale_size_manual(values=rep(0.5,50), guide=F)+
  theme_bw()+
scale_y_continuous(trans="log2")+
 scale_colour_manual(values=c("darkviolet", "cornflower blue", "red3"))+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ylab("Normalised Expression") +
#  xlab('pre')+
  facet_wrap(~dir)




ggplot(mnmd, aes(x=length, y=normexp))+
  geom_point(aes(colour=type))+
#  geom_text(aes(label=sample, colour=type))+
  geom_smooth(method="lm", se=F, aes(size=sname),alpha=0.3,linetype=1, colour="gray30")+
  scale_size_manual(values=rep(0.5,50), guide=F)+
  theme_bw()+
scale_y_continuous(trans="log2")+
 scale_colour_manual(values=c("darkviolet", "cornflower blue", "red3"))+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ylab("Normalised Expression") +
  xlab('Body length')+
  facet_wrap(~dir)


```


##Old vs young by marker

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

sl <- as.character(sampleinfo$sample[sampleinfo$sample %in% colnames(nmd)])

mmd <- gather(as.data.frame(md), key="sample", value="expression", sl)

mnmd <- gather(nmd, key="sample", value="normexp", sl) %>%
  left_join(sampleinfo, by= "sample") %>%
  left_join(mmd, by="sample")


ggplot(mnmd, aes(x= sname, y=normexp, colour=type))+
#  geom_point()+
  theme_bw()+
  geom_boxplot(outlier.shape=NA)+
  scale_y_continuous(trans="log2")+
  facet_grid(~dir, space="free",scales="free" )+
#  geom_boxplot(outlier.shape = NA)+
  scale_colour_manual(values=c("darkviolet", "cornflower blue", "red3"))+
  scale_fill_manual(values=c("darkviolet", "cornflower blue", "red3"))+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ylab("Normalised Expression") +
  xlab('Marker')
  


````



#Models of age based on marker gene expression

##Building the model:

###Multiple linear regression: 

Complete failure of model due to overfitting. Need a better method!

###Random forest

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

md <- as.data.frame(subset(t(ctallgood.norm), sampleinfo$sex == "F" &
  #              sampleinfo$qual == "ok"& 
                 sampleinfo$exp %in% c("Size-age", "cohorts", "OvY") & 
 #                sampleinfo$ctg>2 & 
                sampleinfo$codeset == "phaw_1" &
                 sampleinfo$type %in% c("O", "Y", "M")& 
  !(sampleinfo$sample %in% c("T","C", "U", "LFPD", "O1a", "SFPA", "LFIB","LFIA","AF", "Coh8"))))
md$sample <- rownames(md)

md <- left_join(md, select(sampleinfo, sample, predage), by="sample") %>%
  filter(!(is.na(predage))) %>%
  column_to_rownames(var="sample")




rfagemodel <- train(
  predage~.,
  tuneLength = 4,
  data = md, method = "ranger",
  trControl = trainControl(method = "cv", number = 5, verboseIter = F)
)

rfagemodel
#print("random forest:")
#summary(rfagemodel)
#plot(rfagemodel)

```

##Model accuracy: age from body length regression


```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
md$rfagepred <- predict(rfagemodel)

md <- md %>%
  rownames_to_column(var="sample") %>%
  left_join(select(sampleinfo, -predage), by="sample")


accrf <- lm(predage~rfagepred, md)


ggplot(md, aes(x=predage, y=rfagepred))+
  geom_point(aes(colour=type))+
  xlab("Age (weeks) from body length") +
  ylab("Age (weeks) from marker gene expression (random forest model)")+
  geom_smooth(method="lm", colour="black")+
  theme_bw()+
 scale_colour_manual(values=c("darkviolet", "cornflower blue", "red3"))+
  geom_text_repel(aes(label=sample,colour=type), size=2.5)+
  ggtitle("Using Marker gene expression to predict age", subtitle=paste("R^2 of correlation between measure from length and from gene expression:", round(summary(accrf)$r.squared, 3)))


```


##Model accuracy: cohort age data

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

  
ggplot(md, aes(x=wkage, y=rfagepred))+
  geom_point(aes(colour=type))+
  xlab("Age (weeks): cohorts") +
  ylab("Age (weeks) from marker gene expression (random forest model)")+
  geom_smooth(method="lm", colour="black")+
  theme_bw()+
 scale_colour_manual(values=c("darkviolet", "cornflower blue", "red3"))+
  geom_text_repel(aes(label=sample,colour=type), size=2.5)+
  ggtitle("Using Marker gene expression to predict age", subtitle=paste("R^2 of correlation between actual age and predicted age based on gene expression with random forest model:", round(summary(lm(as.numeric(wkage)~rfagepred, md))$r.squared, 3)))




save(rfagemodel, file="rfagemodel.rdata")

```

So we can use our marker set to very accurately predict a sample's age. 





