---
title: "Aging effect"
author: "Luke Hayden"
date: "July 26, 2018"
output: html_document
---
#Set-up and data import


```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(dplyr)
library(ggbiplot)
library(tibble)
library(tidyr)
library(caret)
library(RColorBrewer)
library(ggrepel)
library(gtools)
````




```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
setwd("~/Documents/nstringjul18")
load(file="allns_data.rdata")

````



#Old vs Young separation
Old vs young as main groups
##PCA old vs young
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}



log <- sampleinfo$reg == "" & 
  !(sampleinfo$age %in% c("N")) & 
  sampleinfo$sex == "F"&
  sampleinfo$exp %in% c("OvY", "Size-age") & 
  !(sampleinfo$sample %in% c("T","C", "U", "LFPD", "O1a", "SFPA", "LFIB","LFIA","AF", "Coh8")) &
 sampleinfo$over10reads > 100 &
  sampleinfo$ctg > 2 &
  sampleinfo$prep == "Luke"

bcmin <- as.matrix(ctallgood.norm[,log== TRUE])
bcmin.groups <- subset(sampleinfo$type, log==TRUE)

tbc <- as.matrix(t(bcmin))

bc.pca <- prcomp(tbc,center = TRUE,scale. = TRUE) 
bc.groups <- bcmin.groups


ggbiplot(bc.pca, obs.scale = 1, var.scale = 1, ellipse = FALSE, circle = TRUE, var.axes=F, labels=rownames(tbc), groups=bc.groups)+ggtitle("PCA of Old vs Young (females): 50 chosen markers")+
    theme_minimal() +
#  scale_colour_manual(values=c("cornflower blue", "red3", "black"))+
  theme(plot.title=element_text(size=8,face="bold"))+
#  geom_label(label=rownames(tbc), aes(colour=bc.groups))+
  scale_colour_manual(values= c("darkviolet","cornflower blue",  "red2"))


````

##What markers best separate old vs young?
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

md <- t(subset(t(ctallgood.norm), sampleinfo$sex == "F" &
  #              sampleinfo$qual == "ok"& 
                 sampleinfo$exp %in% c("OvY", "Size-age", "cohorts") & 
                 sampleinfo$ctg>2 & 
                 sampleinfo$type %in% c("O", "Y", "M")& 
  !(sampleinfo$sample %in% c("T","C", "U", "LFPD", "O1a", "SFPA", "LFIB","LFIA","AF", "Coh8"))))

si <- subset(sampleinfo, sampleinfo$sample %in% colnames(md))
old <- t(subset(t(md), si$age == "O"))
young <- t(subset(t(md), si$age == "Y"))
#regen <- t(subset(t(ctg1to9), sampleinfo$type == "OR" & sampleinfo$qual != "bad"))





nmd <- as.data.frame(md/rowMeans(md))

nmd$name <- rownames(md)
nmd$sname <- substr(nmd$name, 1,2)

oy <- cbind(old, young)
t.result <-  apply(oy, 1, function (x) t.test(x[1:ncol(old)],x[ncol(old)+1: ncol(oy)]))


nmd$p_value <- unlist(lapply(t.result, function(x) x$p.value))

nmd <- left_join(nmd, geneinf, by="name")






nmd$YvOl2fc <- foldchange2logratio(foldchange(rowMeans(young), rowMeans(old)))


nmd$sname <- factor(nmd$sname, levels=nmd$sname[order(nmd$YvOl2fc)])
nmd$good <- nmd$p_value <0.05
nmd$di <- paste0(ifelse(nmd$dir == "upregulated with age", "up", "down"), ": ", nmd$sname)


#nmd <- as.data.frame(md/rowMeans(md))

ggplot(nmd, aes(y=p_value, x=YvOl2fc, fill=good))  +
  scale_shape_manual(values=c(21,22))+
  geom_point(shape=21, colour="black")+ 
  theme_bw()+
  scale_fill_manual(values=c("darkgrey", "orange"), name="",breaks=c(T,F), labels=c("chosen markers", "other")) +
  scale_colour_manual(values=c("red3", "black"))+
  xlab("Absolute Young vs Old Log2Foldchange") +
  ylab("Old vs young p-value")+
  ggtitle(paste(sum(nmd$good), " markers: new optimised set"))+
    geom_hline(linetype=2, yintercept = 0.05)



````

##Old vs young by marker

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

sl <- as.character(sampleinfo$sample[sampleinfo$sample %in% colnames(nmd)])

mmd <- gather(as.data.frame(md), key="sample", value="expression", sl)

mnmd <- gather(nmd, key="sample", value="normexp", sl) %>%
  left_join(sampleinfo, by= "sample") %>%
  left_join(mmd, by="sample")


ggplot(mnmd, aes(x= sname, y=normexp, fill=type))+
#  geom_point()+
  theme_bw()+
  geom_boxplot(outlier.shape=NA)+
  scale_y_continuous(trans="log2")+
  facet_grid(~dir, space="free",scales="free" )+
#  geom_boxplot(outlier.shape = NA)+
  scale_fill_manual(values=c("darkviolet", "cornflower blue", "red3"))+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ylab("Normalised Expression") +
  xlab('Marker')
  


````

##Expression/length relationship

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}





ggplot(mnmd, aes(x=length, y=normexp))+
  geom_point(aes(fill=type), shape=21, colour="white")+
  geom_smooth(method="lm", aes(colour=dir), alpha=0.3, se=F, size=1)+
  theme_bw()+
 scale_y_continuous(trans="log2")+
  scale_colour_manual(values=c("orangered","forest green"))+
  scale_fill_manual(values=c("darkviolet", "cornflower blue",  "red3"))+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ylab("Normalised Expression") +
  xlab('Body length')+
  facet_wrap(~sname, scales="free_y")





ggplot(mnmd, aes(x=length, y=normexp))+
  geom_point(aes(colour=type))+
#  geom_text(aes(label=sample, colour=type))+
  geom_smooth(method="lm", se=F, aes(size=sname),alpha=0.3,linetype=1, colour="gray30")+
  scale_size_manual(values=rep(0.5,50), guide=F)+
  theme_bw()+
scale_y_continuous(trans="log2")+
 scale_colour_manual(values=c("darkviolet", "cornflower blue", "red3"))+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  ylab("Normalised Expression") +
  xlab('Body length')+
  facet_wrap(~dir)


```

#Model length vs expression
All markers as predictors
Linear regression
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

md <- as.data.frame(subset(t(ctall.norm), sampleinfo$sex == "F" &
  #              sampleinfo$qual == "ok"& 
                 sampleinfo$exp %in% c("Size-age", "cohorts", "OvY") & 
                 sampleinfo$ctg>2 & 
                sampleinfo$codeset == "phaw_1" &
                 sampleinfo$type %in% c("O", "Y", "M")& 
  !(sampleinfo$sample %in% c("T","C", "U", "LFPD", "O1a", "SFPA", "LFIB","LFIA","AF", "Coh8"))))
md$sample <- rownames(md)

md <- left_join(md, select(sampleinfo, sample, length), by="sample") %>%
  filter(!(is.na(length))) %>%
  column_to_rownames(var="sample")


model <- train(
  length~., md,
   method = "lm",
  trControl = trainControl(
    method = "cv", number = 5,
    repeats = 5, verboseIter = TRUE
  )
)

model


```


#Model length vs expression
50 chosen markers:
Linear regression
Random forest
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

md <- as.data.frame(subset(t(ctallgood.norm), sampleinfo$sex == "F" &
  #              sampleinfo$qual == "ok"& 
                 sampleinfo$exp %in% c("Size-age", "cohorts", "OvY") & 
 #                sampleinfo$ctg>2 & 
                sampleinfo$codeset == "phaw_1" &
                 sampleinfo$type %in% c("O", "Y", "M")& 
  !(sampleinfo$sample %in% c("T","C", "U", "LFPD", "O1a", "SFPA", "LFIB","LFIA","AF", "Coh8"))))
md$sample <- rownames(md)

md <- left_join(md, select(sampleinfo, sample, length), by="sample") %>%
  filter(!(is.na(length))) %>%
  column_to_rownames(var="sample")


linmodel <- train(
  length~., md,
   method = "lm",
  trControl = trainControl(
    method = "cv", number = 5,
    repeats = 5, verboseIter = TRUE
  )
)


linmodel <- lm(
  length~., md
)

print("linear regression:")
summary(linmodel)
#plot(lmodel)


(rflinmodel <- train(
  length~.,
  tuneLength = 1,
  data = md, method = "ranger",
  trControl = trainControl(method = "cv", number = 5, verboseIter = TRUE)
))


print("random forest:")
model
plot(rflinmodel)


```

Log transformation of the lengths:

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

md <- as.data.frame(subset(t(ctallgood.norm), sampleinfo$sex == "F" &
  #              sampleinfo$qual == "ok"& 
                 sampleinfo$exp %in% c("Size-age", "cohorts", "OvY") & 
 #                sampleinfo$ctg>2 & 
                sampleinfo$codeset == "phaw_1" &
                 sampleinfo$type %in% c("O", "Y", "M")& 
  !(sampleinfo$sample %in% c("T","C", "U", "LFPD", "O1a", "SFPA", "LFIB","LFIA","AF", "Coh8"))))
md$sample <- rownames(md)

md <- left_join(md, select(sampleinfo, sample, length), by="sample") %>%
  filter(!(is.na(length))) %>%
  column_to_rownames(var="sample")

md$length <- log(md$length)

linlogmodel <- train(
  length~., md,
   method = "lm",
  trControl = trainControl(
    method = "cv", number = 5,
    repeats = 5, verboseIter = TRUE
  )
)

print("linear regression:")
linlogmodel


rflogmodel <- train(
  length~.,
  tuneLength = 4,
  data = md, method = "ranger",
  trControl = trainControl(method = "cv", number = 5, verboseIter = TRUE)
)

print("random forest:")
rfmodel
plot(rflogmodel)

md$length <-exp(md$length)
```

Square of the lengths:

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
md$length <- md$length^2


linsqmodel <- train(
  length~., md,
   method = "lm",
  trControl = trainControl(
    method = "cv", number = 5,
    repeats = 5, verboseIter = TRUE
  )
)

print("linear regression:")
linsqmodel


rfsqmodel <- train(
  length~.,
  tuneLength = 4,
  data = md, method = "ranger",
  trControl = trainControl(method = "cv", number = 5, verboseIter = TRUE)
)

print("random forest:")
rfmodel
plot(rfsqmodel)

md$linpred <- predict(linmodel)
md$rflinpred <- predict(rflinmodel)
md$linlogpred <- exp(predict(linlogmodel))
md$rflogpred <-  exp(predict(rflogmodel))
md$linsqpred <- exp(predict(linsqmodel))
md$rfsqpred <-  exp(predict(rfsqmodel))


md$length <- sqrt(md$length)

md <- md %>%
  rownames_to_column(var="sample") %>%
  left_join(select(sampleinfo,wkage, sample) , by=("sample"="sample") )


```


```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}



preddf <- data.frame(length=md$length , 
                     sample = rownames(md), 
                     wkage =md$wkage,

                     rflinpred = predict(rflinmodel),
                     linlogpred = exp(predict(linlogmodel)),
                     rflogpred =  exp(predict(rflogmodel)), 
                     
                     linsqpred = sqrt(predict(linsqmodel)),
                     rfsqpred =  sqrt(predict(rfsqmodel))
                     
)


mpreddf <- gather(preddf, key="model", value="prediction", -length, -sample, -wkage)




ggplot(mpreddf, aes(x=length, y=prediction, colour=model))+
  geom_point()+
  geom_point(aes(y=length), colour="black")+
  facet_wrap(~model, ncol=2)+
  theme_bw()



```


#Models of age based on marker gene expression

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

md <- as.data.frame(subset(t(ctallgood.norm), sampleinfo$sex == "F" &
  #              sampleinfo$qual == "ok"& 
                 sampleinfo$exp %in% c("Size-age", "cohorts", "OvY") & 
 #                sampleinfo$ctg>2 & 
                sampleinfo$codeset == "phaw_1" &
                 sampleinfo$type %in% c("O", "Y", "M")& 
  !(sampleinfo$sample %in% c("T","C", "U", "LFPD", "O1a", "SFPA", "LFIB","LFIA","AF", "Coh8"))))
md$sample <- rownames(md)

md <- left_join(md, select(sampleinfo, sample, predage), by="sample") %>%
  filter(!(is.na(predage))) %>%
  column_to_rownames(var="sample")




rfagemodel <- train(
  predage~.,
  tuneLength = 4,
  data = md, method = "ranger",
  trControl = trainControl(method = "cv", number = 5, verboseIter = TRUE)
)

print("random forest:")
rfmodel
plot(rfagemodel)


md$rfagepred <- predict(rfagemodel)
md <- md %>%
  rownames_to_column(var="sample") %>%
  left_join(select(sampleinfo, -predage), by="sample")


acc <- lm(predage~rfagepred, md)


ggplot(md, aes(x=predage, y=rfagepred))+
  geom_point(aes(colour=type))+
  xlab("Age (weeks) from body length") +
  ylab("Age (weeks) from marker gene expression (random forest model)")+
  geom_smooth(method="lm", colour="black")+
  theme_bw()+
 scale_colour_manual(values=c("darkviolet", "cornflower blue", "red3"))+
  geom_text_repel(aes(label=sample,colour=type), size=2.5)+
  ggtitle("Using Marker gene expression to predict age", subtitle=paste("R^2 of correlation between measure from length and from gene expression:", round(summary(acc)$r.squared, 3)))



  
ggplot(md, aes(x=wkage, y=rfagepred))+
  geom_point(aes(colour=type))+
  xlab("Age (weeks): cohorts") +
  ylab("Age (weeks) from marker gene expression (random forest model)")+
  geom_smooth(method="lm", colour="black")+
  theme_bw()+
 scale_colour_manual(values=c("darkviolet", "cornflower blue", "red3"))+
  geom_text_repel(aes(label=sample,colour=type), size=2.5)+
  ggtitle("Using Marker gene expression to predict age", subtitle=paste("R^2 of correlation between measure from length and from gene expression:", round(summary(lm(as.numeric(wkage)~rfagepred, md))$r.squared, 3)))




save(rfagemodel, file="rfagemodel.rdata")

```

So we can use our marker set to very accurately predict a sample's age. 





