---
title: "PCAfull"
author: "Luke Hayden"
date: "1 October 2019"
output: html_document
---



```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(dplyr)
library(ggbiplot)
library(tibble)
library(tidyr)
library(reshape2)
library(matrixStats)
library(gtools)
library(GGally)
library(philentropy)
library(ggrepel)
library(Rmisc)
```
#PCA all
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

sampleinfob <- read.csv("sialplus.csv")



sampleinfoc <- subset(sampleinfob, sampleinfob$sample %in% colnames(ctfull.norm))



md <- as.data.frame(subset(t(ctfull.norm), 
                sampleinfoc$sex != "M" &              
#                  sampleinfoc$prep== "Luke"&
       #           sampleinfoc$
                 sampleinfoc$exp %in% c("Size-age", "cohorts", "OvY") &
                !(is.na(sampleinfoc$length)) &
                 sampleinfoc$type %in% c("O", "Y", "M")
))

md <-subset(md, rownames(md)%in% qualsum$sample[qualsum$good])
#predage <- predict(polymod3, newdata=sampleinfoc)

#sampleinfoc$predage <- predage

log <-     sampleinfob$sample %in% rownames(md)

bcmin <- as.matrix(ctfull.norm[,log==T]) 
bcmin <- bcmin[  complete.cases(bcmin),]
bcmin.groups <- sampleinfoc$type[log]
tbc <- as.matrix(t(bcmin))
bc.pca <- prcomp(tbc,center = TRUE,scale. = TRUE) 
bc.groups <- bcmin.groups


ggbiplot(bc.pca, obs.scale = 1, var.scale = 1, ellipse = FALSE, circle = TRUE, var.axes=F, labels=rownames(tbc), groups=bc.groups)+
  ggtitle("PCA based on housekeeping gene expression")+
    theme_minimal() 


```
#  scale_colour_manual(values=c("cornflower blue", "red3", "black"), guide=F)

pcat <- bc.pca$x %>%
  as.data.frame() %>%
  rownames_to_column(var="sample") %>%
  left_join(sampleinfoc, by="sample")

spca <- summary(bc.pca)$importance


(pcalm <- summary(lm(data=pcat, predage~PC1)))

(
  p= ggplot(pcat, aes(y=PC1, x=predage,label=sample))+
  geom_point()+
  geom_smooth(method= "lm", se=F)+
  theme_bw()+
  xlab("Age (weeks) from body length") +
  ylab(paste0("PC1 (", round(spca[2,1] *100, 1), "% of variation) \nfrom marker gene expression"))+
  ggtitle("PCA to divide old vs young", subtitle = paste0("R-squared: ", round(pcalm$r.squared, 3)))+
  theme(
    plot.background = element_rect(fill = "transparent",colour = "transparent") # bg of the panel
  )

)


ggsave(plot=p,height=8,width=12,dpi=300, device= "png", filename=paste("pca.png"),  bg = "transparent",limitsize = FALSE)


```


#PCA train test
```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

sampleinfob <- read.csv("sialplus.csv")



sampleinfoc <- subset(sampleinfob, sampleinfob$sample %in% colnames(ctfull.norm))


predage <- predict(polymod3, newdata=sampleinfoc)

sampleinfoc$predage <- predage

md <- as.data.frame(subset(t(ctfull.norm), 
                sampleinfoc$sex == "F" &              
                  sampleinfoc$prep== "Luke"&
       #           sampleinfoc$
                 sampleinfoc$exp %in% c("Size-age", "cohorts", "OvY") &
                !(is.na(sampleinfoc$length)) &
                 sampleinfoc$type %in% c("O", "Y", "M")
))


trainchoice <- sample(1:nrow(md))[1:ceiling(4*(nrow(md)/5))]

md <- md[  complete.cases(md),]
trdat<- md[trainchoice,]
tedat<- md[-trainchoice,]


log <-     sampleinfob$sample %in% rownames(trdat)

bcmin <- as.matrix(ctfull.norm[,log==T]) 

bcmin.groups <- sampleinfoc$type[log]
tbc <- as.matrix(t(bcmin))
bc.pca <- prcomp(tbc,center = TRUE,scale. = TRUE) 
bc.groups <- bcmin.groups


ggbiplot(bc.pca, obs.scale = 1, var.scale = 1, ellipse = FALSE, circle = TRUE, var.axes=F, labels=rownames(tbc), groups=bc.groups)+
  ggtitle("PCA based on housekeeping gene expression")+
    theme_minimal() 
#  scale_colour_manual(values=c("cornflower blue", "red3", "black"), guide=F)



tedat.sc <- scale(tedat, center= bc.pca$center)
tedat.pred <- tedat.sc %*% bc.pca$rotation 
tetrpca <- rbind(bc.pca$x, tedat.pred)
                   
                   



pcat <- tetrpca %>%
  as.data.frame() %>%
  rownames_to_column(var="sample") %>%
  left_join(sampleinfoc, by="sample")

pcat$intrain <- pcat$sample %in% rownames(trdat)

spca <- summary(bc.pca)$importance


(pcalm <- summary(lm(data=pcat, predage~PC1)))

(
  p= ggplot(pcat, aes(y=PC1, x=predage,label=sample, colour=intrain))+
  geom_point()+
  geom_smooth(method= "lm", se=F, inherit.aes = F, aes(y=PC1, x=predage), colour="black")+
  theme_bw()+
 scale_colour_manual(values=c("cornflower blue", "red3"), name="Data partition", labels=c("Test", "Training"))+
  xlab("Age (weeks) from body length") +
  ylab(paste0("PC1 (", round(spca[2,1] *100, 1), "% of variation) \nfrom marker gene expression"))+
  ggtitle("PCA to divide old vs young", subtitle = paste0("R-squared: ", round(pcalm$r.squared, 3)))+
  theme(
    plot.background = element_rect(fill = "transparent",colour = "transparent") # bg of the panel
  )

)


ggsave(plot=p,height=8,width=12,dpi=300, device= "png", filename=paste("pcatrte.png"),  bg = "transparent",limitsize = FALSE)


```