---
title: "PCA vs LDA"
author: "Luke Hayden"
date: "29th March 2019"
output: html_document
---


```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(dplyr)
library(ggbiplot)
library(tibble)
library(tidyr)
library(caret)
library(RColorBrewer)
library(ggrepel)
library(gtools)
library(MASS)
````




```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
#setwd("~/Documents/nstringjul18")
load(file="allns_data.rdata")
load(file="markerchoiceinfo.rdata")

load(file="qualitydat.rdata")
load(file="sampleinfo.rdata")

````



Old vs young as main groups

#PCA old vs young
For comparative purposes


```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}



log <- sampleinfo$reg == "" & 
  !(sampleinfo$age %in% c("N", "M")) & 
  sampleinfo$sample %in% qualsum$sample[qualsum$good] & 
  sampleinfo$sex == "F"&
  sampleinfo$exp %in% c("OvY", "Size-age") & 
  sampleinfo$ctg > 2 &
  sampleinfo$prep == "Luke"

bcmin <- as.matrix(ctall.norm[,log== TRUE])
bcmin.groups <- subset(sampleinfo$type, log==TRUE)

tbc <- as.matrix(t(bcmin))

bc.pca <- prcomp(tbc,center = TRUE,scale. = TRUE) 
bc.groups <- bcmin.groups


ggbiplot(bc.pca, obs.scale = 1, var.scale = 1, ellipse = FALSE, circle = TRUE, var.axes=F, labels=rownames(tbc), groups=bc.groups)+ggtitle("PCA of Old vs Young (females): 195 markers")+
    theme_bw() +
  scale_colour_manual(values= c("cornflower blue",  "red2"))


````

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}



log <- sampleinfo$reg == "" & 
  !(sampleinfo$age %in% c("N", "M")) & 
  sampleinfo$sample %in% qualsum$sample[qualsum$good] & 
  sampleinfo$sex == "F"&
  sampleinfo$exp %in% c("OvY", "Size-age") & 
  sampleinfo$ctg > 2 &
  sampleinfo$prep == "Luke"

bcmin <- as.matrix(ctallgood.norm[,log== TRUE])
bcmin.groups <- subset(sampleinfo$type, log==TRUE)

tbc <- as.matrix(t(bcmin))

bc.pca <- prcomp(tbc,center = TRUE,scale. = TRUE) 
bc.groups <- bcmin.groups


ggbiplot(bc.pca, obs.scale = 1, var.scale = 1, ellipse = FALSE, circle = TRUE, var.axes=F, labels=rownames(tbc), groups=bc.groups)+ggtitle("PCA of Old vs Young (females): 50 chosen markers")+
    theme_bw() +
  scale_colour_manual(values= c("cornflower blue",  "red2"))


````



##LDA

Like PCA, this is also a dimensionality reduction technique and a type of unsupervised machine learning. Given samples from two classes and their values obtained for a set of features (our markers), this method looks to find the axis that best separates these classes, giving each feature a loading. I have run this for the full marker set and for the reduced set. The method works out a way of dividing the classes (linear weighting of each marker's value) and outputs a position along this axis for each sample. 

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
log <- sampleinfo$reg == "" & 
  !(sampleinfo$age %in% c("N", "M")) & 
  sampleinfo$sample %in% qualsum$sample[qualsum$good] & 
  sampleinfo$sex == "F"&
  sampleinfo$exp %in% c("OvY", "Size-age") & 
  sampleinfo$ctg > 2 &
  sampleinfo$prep == "Luke"

bcmin <- t(as.matrix(ctall.norm[,log== TRUE])) %>%
  as.data.frame() %>%
  rownames_to_column(var= "sample") %>%
  left_join(dplyr::select(sampleinfo, sample, type), by= "sample") %>%
  column_to_rownames("sample")

bcmin$type <- factor(bcmin$type)

lda <- lda(formula = type ~ ., 
         data = bcmin, 
         prior = c(1,1)/2)#, CV = TRUE)

lp <- predict(object=lda, newdata=dplyr::select(bcmin, -type))



lpp <- lp$posterior %>%
  as.data.frame() %>%
  rownames_to_column(var= "sample") %>%
  left_join(sampleinfo, by= "sample")

lpp$ldaclass <- lp$class
lpp$iscorrect <- as.character(lpp$type) == as.character(lpp$ldaclass)
  

ggplot(lpp, aes(x=O, fill=type))+
  geom_histogram()+
  ggtitle("LDA of Old vs Young (females): all 195 markers")+
    theme_bw() +
  scale_fill_manual(values= c("cornflower blue",  "red2"), name= "Age category", labels = c("Old", "Young"))+
  xlab("Old/Young classification posterior probability")


ggplot(lpp, aes(y=O,x=type, color=type, shape=iscorrect))+
  geom_point(size=3, alpha=0.5)+
  ggtitle("PCA of Old vs Young (females): all 195 markers")+
    theme_bw() +
  scale_color_manual(values= c("cornflower blue",  "red2"), name= "Age category", labels = c("Old", "Young"))+
  scale_shape_manual(values= c(4,16), name= "Accuracy", labels= c("Incorrect", "Correct"))+
  xlab("Age (weeks) from body length") +
  ylab("Old/Young classification posterior probability")
````


```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
log <- sampleinfo$reg == "" & 
  !(sampleinfo$age %in% c("N", "M")) & 
  sampleinfo$sample %in% qualsum$sample[qualsum$good] & 
  sampleinfo$sex == "F"&
  sampleinfo$exp %in% c("OvY", "Size-age") & 
  sampleinfo$ctg > 2 &
  sampleinfo$prep == "Luke"

bcmin <- t(as.matrix(ctallgood.norm[,log== TRUE])) %>%
  as.data.frame() %>%
  rownames_to_column(var= "sample") %>%
  left_join(dplyr::select(sampleinfo, sample, type), by= "sample") %>%
  column_to_rownames("sample")

bcmin$type <- factor(bcmin$type)

lda <- lda(formula = type ~ ., 
         data = bcmin, 
         prior = c(1,1)/2)#, CV = TRUE)

lp <- predict(object=lda, newdata=dplyr::select(bcmin, -type))



lpp <- lp$posterior %>%
  as.data.frame() %>%
  rownames_to_column(var= "sample") %>%
  left_join(sampleinfo, by= "sample")

lpp$ldaclass <- lp$class
lpp$iscorrect <- as.character(lpp$type) == as.character(lpp$ldaclass)
  


ggplot(lpp, aes(x=O, fill=type))+
  geom_histogram()+
  ggtitle("LDA of Old vs Young (females): 50 markers")+
    theme_bw() +
  scale_fill_manual(values= c("cornflower blue",  "red2"), name= "Age category", labels = c("Old", "Young"))+
  xlab("Old/Young classification posterior probability")


ggplot(lpp, aes(y=O,x=type, color=type, shape=iscorrect))+
  geom_point(size=3, alpha=0.5)+
  ggtitle("PCA of Old vs Young (females): 50 markers")+
    theme_bw() +
  scale_color_manual(values= c("cornflower blue",  "red2"), name= "Age category", labels = c("Old", "Young"))+
  scale_shape_manual(values= c(4,16), name= "Accuracy", labels= c("Incorrect", "Correct"))+
  xlab("Age (weeks) from body length") +
  ylab("Old/Young classification posterior probability")
````


##Perspectives

LDA works fairly well, might be a nice panel to put into the paper as illustrative of simpler methods than the ensemble models. Might also have its place in the pipeline in marker choice for models. 